---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mapr

**mapr** is an R package that makes it easier to make maps in R  

Given a set of locations, for example from a tagged marine animal, the `mapr` function will load a global shapefile from the Natural Earth database using `rworldmap` and manipulate it for plotting using `ggplot2` and `sf`

Alternatively, given the same set of locations the `meshr` function will help you to create a shapefile (or nested list of shapefiles) that can be used as a boundary when creating an `INLA` mesh using `inla.mesh.2d`

Given a set of locations, for example from a tagged marine animal, the `mapr` function will load a global shapefile from the Natural Earth database using `rworldmap` and manipulate it for plotting using `ggplot2`

Alternatively, given the same set of locations the `meshr` function will help you to create a shapefile that can be used as a boundary when creating an `INLA` mesh using `inla.mesh.2d`

## Installation

To download the current development version from GitHub:

```{r install, message = F, warning = F}
# install.packages("devtools")  
devtools::install_github("jamesgrecian/mapr")
```


### An example map

Here's an example of how to generate a map containing a coastline and some animal locations using the `mapr` function alongside the `sf` and `ggplot2` libraries

```{r mapr example with ellies, message = F, warning = F}
# load libraries
require(tidyverse)
require(sf)
require(mapr)

# load example dataset
data(ellie)

# define an appropriate proj.4 projection
prj <- '+proj=laea +lat_0=-60 +lon_0=70 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs'

# use mapr to generate a shapefile
world_shp <- mapr(ellie, prj, buff = 1e6)

# output a plot using ggplot
p1 <- ggplot() +
  geom_sf(aes(), data = world_shp) +
  geom_sf(aes(), data = st_as_sf(ellie, coords = c('lon', 'lat'))
          %>% st_set_crs('+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs'))

print(p1)

```


## An example INLA mesh

When using INLA to analyse animal movement data it is useful to base the mesh on the distribution of locations.  

Here's an example of how to generate a boundary shapefile using `meshr` that can then be passed to `inla.mesh.2d`, we can plot the mesh using `sf` and `ggplot2`

```{r meshr example with ellies, message = F, warning = F}
# load libraries
require(tidyverse)
require(sf)
require(mapr)
require(INLA)
require(inlabru)

# load example dataset
data(ellie)

# define an appropriate proj.4 projection
prj <- '+proj=laea +lat_0=-60 +lon_0=70 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs'

# use meshr to generate the boundary
b <- meshr(ellie, prj, buff = 5e5, keep = 0.05, Neumann = T)

# use the boundary to generate the INLA mesh
mesh = inla.mesh.2d(boundary = b, max.edge = c(250000, 1e+06), cutoff = 25000, max.n = 500)

# output a plot using ggplot
p2 <- ggplot() + 
  geom_sf(aes(), data = mapr(ellie, prj, buff = 1e6)) +
  inlabru::gg(mesh) +
  geom_sf(aes(), data = st_as_sf(ellie, coords = c("lon", "lat"))
          %>% st_set_crs("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"))

print(p2)

```
