---
title: "A guide to making maps in R using the mapr package"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{A guide to making maps in R using the mapr package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  out.width = "100%",
  dpi = 150
)
```

It's easy to make a quick map in R using just a few steps. It can be tricky to
make a publication ready map consistently. That's where `mapr` can help. Given a
set of locations, for example from a tagged marine animal, the main `mapr`
function will load a global shapefile from the Natural Earth database and
automagically manipulate it for plotting using `ggplot2` and `sf`.

This guide builds a map of seabird tracks one step at a time, starting with the
plot most people make first and ending with one you could put in a report. Each
step changes one thing, so you can see what each piece of code does.

We'll use three packages: `mapr`, `sf` and `ggplot2`. There is a small example
dataset within `mapr` containing GPS tracks from three immature northern gannets
tagged on the Bass Rock.

```{r packages, message = FALSE}
library(mapr)
library(sf)
library(ggplot2)

data(gannets)
head(gannets)
```

The tracks are in a `tibble`, which is a form of "tidy" data frame, with
longitude and latitude columns.

## 1. The first attempt

The obvious start is a coastline with the tracks plotted on top as points,
longitude on x and latitude on y.

The `rnaturalearth` R package provides an easy way to download a shapefile and
plot a map of the world. If you ask for the `sf` version then plotting with
`geom_sf()` makes dealing with spatial data a lot easier.

```{r naive-world}
world_shp <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

ggplot() +
  geom_sf(data = world_shp) +
  geom_point(data = gannets, aes(x = lon, y = lat),
             colour = "firebrick", size = 0.1)
```

That's the whole world, with the gannets as a red smudge off Scotland. Natural
Earth gives you every country, and ggplot draws everything it's given.

## 2. Zooming in

The natural next move is to zoom in by setting the limits, in degrees:

```{r naive-zoom}
ggplot() +
  geom_sf(data = world_shp) +
  geom_point(data = gannets, aes(x = lon, y = lat),
             colour = "firebrick", size = 0.1) +
  coord_sf(xlim = c(-8, 12), ylim = c(50, 63))
```

This looks like a North Sea map, and for a quick look it's fine. But look at the
grid lines. The lines of longitude are parallel, so every grid cell is a
rectangle. On the real Earth, lines of longitude converge towards the pole: a
degree of longitude is about 111 km wide at the equator and only about 55 km at
Shetland. The map is quietly stretching the north and squashing the south.

## 3. Projecting

The fix is to draw the map in a *projection*: a way of flattening the curved
Earth that keeps distances and shapes close to right over the area you care
about. For the North Sea, a common choice is UTM zone 30N:

```{r prj}
prj <- "+proj=utm +zone=30 +datum=WGS84 +units=m +no_defs"
```

So far ggplot has treated `lon` and `lat` as ordinary x and y numbers. That
worked because the coastline was in longitude and latitude too, so the numbers
matched. Once the map is projected, x and y are metres, and plotting raw
longitudes as if they were metres would put the gannets somewhere near the
equator. ggplot needs to know the tracks are coordinates, and which coordinate
system they're in, so it can convert them.

That's what `st_as_sf()` does. GPS data is almost always longitude and latitude
on WGS84, which has the code 4326:

```{r tracks}
gannets_sf <- gannets |> st_as_sf(coords = c("lon", "lat"), crs = 4326)
```

Note the order in `coords`: longitude first, then latitude. Get it the wrong way
round and your gannets will turn up in the Indian Ocean.

Now that we've told R the gannet data is spatial we can plot it natively in
ggplot using `geom_sf()`:

```{r first-sf-plot}
ggplot() +
  geom_sf(data = world_shp) +
  geom_sf(data = gannets_sf, colour = "firebrick", size = 0.1) +
  coord_sf(xlim = c(-8, 12), ylim = c(50, 63))
```

But although the data are spatial they have not yet been projected. We can do
this with `st_transform()`:

```{r projecting-data}
world_utm <- world_shp |> st_transform(prj)
gannets_utm <- gannets_sf |> st_transform(prj)

ggplot() +
  geom_sf(data = world_utm) +
  geom_sf(data = gannets_utm, colour = "firebrick", size = 0.1)
```

It is obviously wrong to project the whole world onto a UTM30N projection, but
it shows you how the projection works: it is correct for our study area and
increasingly wrong everywhere else. So instead we crop. Note that the limits are
now in metres, not degrees, and that the graticules now curve.

```{r zoom-in-projection}
ggplot() +
  geom_sf(data = world_utm) +
  geom_sf(data = gannets_utm, colour = "firebrick", size = 0.1) +
  coord_sf(xlim = c(300000, 1200000), ylim = c(5700000, 6800000))
```

If you're wondering where numbers like those come from, `st_bbox(gannets_utm)`
gives the extent of the tracks in metres, which is a good starting point.

## 4. Letting mapr do the work

At this point you can begin to assemble a plot for publication. R and ggplot
do a really good job of making it intuitive, but you can spend ages tweaking
the crop and layout. So at this point I'd introduce you to `mapr`, a small R
package I wrote to help.

`mapr()` does three jobs at once: it fetches the shapefile we can use for the
coastline, crops it to the area around your tracks, and projects it to a
user defined projection. The `buff` argument says how far beyond the tracks you
want the file to go, in metres. ggplot automatically pads a plot by 5% on each
side; setting `expand = FALSE` means you don't get empty space around your land
shapefile.

```{r mapr}
land <- mapr(gannets, prj, buff = 400000)

ggplot() +
  geom_sf(data = land) +
  geom_sf(data = gannets_utm, colour = "firebrick", size = 0.1) +
  coord_sf(crs = prj, expand = FALSE) 
```

This is the same projected map as the one above, but `mapr()` has worked out the
crop from the tracks for you, and only the land you need has been kept.

## 5. Framing

The map above stops wherever `mapr()` happened to crop it, and the edges are cut
straight through the land. The larger the area you are studying, the further
from a square shape your graticules will be. For example, polar projections
often result in circular plots.

`make_map_furniture()` is designed to help give a projected map a proper frame.
You choose the extent in degrees, and how often you want grid lines:

```{r furniture}
mf <- make_map_furniture(
  xmin = -7.5, xmax = 10, ymin = 50, ymax = 62,
  crs = prj,
  buffer = 200000,
  lat_by = 5,
  lon_by = 5
)
```

The extent here defines the plot frame in longitude and latitude. The `buffer`
is a margin around the frame, in metres, which we'll use for labels in step 6.

The function returns a list of pieces to add to the plot:

- `mf$cookie` is a white shape with a hole the size of your map. Drawn on top of
  everything else, it hides anything that spills past the frame.
- `mf$flyway` is the outline of the frame itself.
- `mf$xlim` and `mf$ylim` are the plot limits, in the projection's units.
- `mf$meridians` and `mf$parallels` are the grid positions and their labels.

The land from step 4 needs to reach past the frame on every side. The cookie
hides anything beyond the frame, but it can't add land that isn't there, so if
you see a straight edge cutting through the land inside the frame, increase
`buff` in `mapr()`.

```{r framed}
ggplot() +
  theme_minimal(base_size = 8) +
  geom_sf(data = land) +
  geom_sf(data = gannets_utm, colour = "firebrick", size = 0.1) +
  geom_sf(data = mf$cookie, fill = "white", colour = NA) +
  geom_sf(data = mf$flyway, fill = NA) +
  scale_x_continuous(breaks = mf$meridians$lon) +
  scale_y_continuous(breaks = mf$parallels$lat) +
  coord_sf(xlim = mf$xlim, ylim = mf$ylim, crs = prj, expand = FALSE)
```

The order of the layers matters. Land and tracks go first, then the cookie over
the top of them, then the frame. Put the cookie first and it will be hidden
under the land.

The two `scale_` lines set where the grid lines go, using the positions from the
furniture, so the grid matches the extent you asked for. Without them ggplot
picks its own spacing.

ggplot has also automatically put the axis labels around the buffer area,
instead we can replace these with the ones stored in the map furniture.

## 6. Labels

ggplot labels the grid lines itself, at the edge of the plot. With a small
buffer and a near-rectangular frame that can work well enough, but with a wide
buffer like this one the labels end up out at the edge of the margin, away from
the frame. On a map where the frame is noticeably curved they also end up in odd
places. It's better to put them on the frame yourself:

```{r labels}
ggplot() +
  theme_minimal(base_size = 8) +
  geom_sf(data = land) +
  geom_sf(data = gannets_utm, colour = "firebrick", size = 0.1) +
  geom_sf(data = mf$cookie, fill = "white", colour = NA) +
  geom_sf(data = mf$flyway, fill = NA) +
  scale_x_continuous(breaks = mf$meridians$lon) +
  scale_y_continuous(breaks = mf$parallels$lat) +
  coord_sf(xlim = mf$xlim, ylim = mf$ylim,
           crs = prj, expand = FALSE) +
  theme(axis.text = element_blank(),
        axis.title = element_blank()) +
  geom_sf_text(data = mf$parallels, aes(label = label),
               size = 2.5, colour = "grey40",
               nudge_x = -15000, hjust = 1) +
  geom_sf_text(data = mf$meridians, aes(label = label),
               size = 2.5, colour = "grey40",
               nudge_y = -15000, vjust = 1)
```

What changed:

- `axis.text = element_blank()` hides ggplot's own labels, so they don't appear
  twice.
- The two `geom_sf_text()` layers put the furniture's labels just outside the
  frame. `nudge_x` and `nudge_y` move them out by 15 km, and `hjust` and
  `vjust` line them up against that point.

If the labels sit too close or too far from the frame, change the nudge. If they
don't fit at all, make the buffer bigger.

## 7. Finishing

The rest is ordinary ggplot. Here the land is a quieter grey, each bird has its
own colour, and the colony is marked:

```{r colony}
colony <- st_as_sf(data.frame(lon = -2.64, lat = 56.08),
                   coords = c("lon", "lat"), crs = 4326)
```

```{r final}
p <- ggplot() +
  theme_minimal(base_size = 8) +
  geom_sf(data = land, fill = "grey85", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = gannets_utm, aes(colour = id), size = 0.1, show.legend = FALSE) +
  geom_sf(data = colony, shape = 24, size = 2, fill = "gold", colour = "black") +
  geom_sf(data = mf$cookie, fill = "white", colour = NA) +
  geom_sf(data = mf$flyway, fill = NA, linewidth = 0.3) +
  scale_x_continuous(breaks = mf$meridians$lon) +
  scale_y_continuous(breaks = mf$parallels$lat) +
  scale_colour_brewer(palette = "Dark2") +
  coord_sf(xlim = mf$xlim, ylim = mf$ylim,
           crs = prj, expand = FALSE) +
  labs(title = "Immature gannets from the Bass Rock, 2015") +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_line(colour = "grey90", linewidth = 0.2),
        plot.title = element_text(face = "bold", hjust = 0.5)) +
  geom_sf_text(data = mf$parallels, aes(label = label),
               size = 2.5, colour = "grey40",
               nudge_x = -15000, hjust = 1) +
  geom_sf_text(data = mf$meridians, aes(label = label),
               size = 2.5, colour = "grey40",
               nudge_y = -15000, vjust = 1)

p
```

To save it, give `ggsave()` a size and a resolution. 300 dpi is the usual
minimum for print. Setting the size to what you need the figure to be on the
page means text will be printed at the correct font size.

```{r save, eval = FALSE}
ggsave("gannets_map.png", p, width = 10, height = 15, units = "cm", dpi = 300)
```

As a side note, because the map is a ggplot, it's easy to add complexity, for
example by faceting:

```{r facet}
p + facet_wrap(~id)
```
